---
title: "RBP4 and dementia"
author: "Chris McNeil"
date: "19/04/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Is Retinol Binding Protein 4 plasma contentration associated with cognitive performance and markers of AD in a normal older people?

## Abstract
Retinol Binding protein has been linked to cognitive decline and Alzheimer's disease, both in humans, and in animal models of AD. Here we measured the plasma concentration of RBP4 in the participants of the Stratifying Depression Longitudinally (STRADL) cohort. 


**Hypothesis - RBP4 plasma concentration associated with cognitive decline and risk of developing Alzheimer's dementia**

* Plasma analysis of RBP4 by Jason Clarke, under supervision of Prof Peter McCafferty.
* This analysis by Chris McNeil. 
* 2021_02_24

## Data Analysis

**Import Packages**
```{r  packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(readr)
library(readxl)
library(performance)
library(Hmisc) #correlations
library(corrplot) #plotting correlations
library(expss) # apply labels to variable names
library(knitr) #neat tables
library(kableExtra) #neater tables
library(viridis)
library(RColorBrewer)
library(moderndive) # linear regression
```

### Prepare Data
* import STRADL data
  + Confounding variables: demographics
  + Variables of interest: Brain MRI variables (volumes etc), cognitive variables
```{r data_import}
nonmridata <- read_excel("rawdata/2021_02_25_Cmcneil.xlsx")
RBP4_data <- read_excel("rawdata/RBP4 data 1.xlsx", 
    col_types = c("skip", "text", "skip", 
        "numeric", "numeric"))
mri_data <- read_excel("rawdata/mri_data.xlsx")
```

**Rename variables
```{r renaming}
nonmridata <- nonmridata %>% rename(id=ID)
mri_data <- mri_data %>% rename(id=STRADLID)
RBP4_data <- RBP4_data %>% rename(id=`SAMPLE CODE`)
```

**Merge dataframes
```{r merging, echo=FALSE}
main_data<-left_join(RBP4_data,mri_data)
main_data <- left_join(main_data,nonmridata)

rm(list=setdiff(ls(), "main_data"))
```

**Calculate volumes as ml to clarify**
```{r}
main_data <- main_data %>% mutate(BrainSegNotVent=BrainSegNotVent/100,CortexVol=CortexVol/1000, LeftHippocampus=LeftHippocampus/1000,rightHippocampusVol=rightHippocampusVol/1000)
```

## Calculate relative brain volumes

```{r}
main_data <- main_data %>% mutate(total_hippo=LeftHippocampus+rightHippocampusVol)
main_data <- main_data %>% mutate(hippo_brain=total_hippo/BrainSegNotVent,hippo_icv=total_hippo/eTIV,hippo_brain=total_hippo/BrainSegNotVent,brain_ICV=BrainSegNotVent/eTIV, lh_brain=LeftHippocampus/brain_ICV,rh_brain= rightHippocampusVol/BrainSegNotVent,lh_icv=LeftHippocampus/eTIV, rh_icv=rightHippocampusVol/eTIV)

```

**change sex character to numeric code (0=female, 1=male). NB specify to use dplyr because of masking by expss library.**
```{r recode}
main_data <- main_data %>% mutate(Sex=dplyr::recode(Sex,"M"=1,"F"=0))
```

```{r}
#calculate BMI
main_data <- main_data %>% mutate(BMI=`Weight(kg)`/((`Height(cm)`*`Height(cm)`)/10000))
```

**label  variables to make plots/tables easier to read:**
```{r}
main_data  <- main_data %>% apply_labels(id="ID",
                                         BrainSegNotVent = "Brain Vol (ml)",
                                         CortexVol='Cortex Vol (ml)',
                                         LeftHippocampus = "L Hippo (ml)",
                                         rightHippocampusVol = "R Hippo (ml)",
                                         eTIV = "ICV (ml)",
                                         digsym='Digit Score',
                                         AgeFaceToFace='Age (yrs)',
                                         mema='Memory (immediate)',
                                         memdela ='Memory (delayed)',
                                         vftot='Verbal Fluency',
                                         mrtotc='Matrix Reasoning (correct)',
                                         total_hippo='total Hippo (ml)',
                                        hippo_icv='TH/IVC')
```


#### Data normalising and transforming units
* Check data distribution visually
    + attempt to fit a normal distribution curve to data

```{r distribution}
main_data %>% ggplot(aes(`RBP4 (ug/ml)`)) + geom_histogram(binwidth = 10,color="black", fill = "steelblue", alpha=.4,aes( y=..density..))+stat_function(fun = dnorm, args = list(mean =mean(main_data$`RBP4 (ug/ml)`), sd=sd(main_data$`RBP4 (ug/ml)`))) + theme_bw()

```

**Transform the RBP4 data using log10**
```{r}
main_data <- main_data %>% mutate(RBP4_log10=log10(`RBP4 (ug/ml)`))
```
**Check distribution**
```{r}
main_data %>% ggplot(aes(RBP4_log10)) + geom_histogram(bins = 35,color="black", fill = "steelblue", alpha=.4,aes( y=..density..))+stat_function(fun = dnorm, args = list(mean =mean(main_data$RBP4_log10), sd=sd(main_data$RBP4_log10))) + theme_bw()
```

### Data summary

**Summarise data by sex**
```{r fig.cap="Plasma ug.ml-1"}
not.na = function(x) {sum(!is.na(x))}
stats = c('mean', 'median', 'sd', 'not.na')
kable(main_data %>%select(`RBP4 (ug/ml)`,Sex) %>% group_by(Sex) %>% summarise_all(.funs=stats),digits = 2) %>%kable_styling(font_size = 16) %>%  kable_minimal(full_width = F, html_font = "Cambria") 


```
**Table 1.** [RBP4]^ug.ml-1^ by sex.


```{r}
plotBMIvsRBP4 <- main_data %>% ggplot(aes(BMI,RBP4_log10))+geom_point()+geom_smooth() + theme_bw()

plotBMIvsRBP4
```
**Figure 1.** [RBP4]^ug.ml-1^ and BMI
**RBP4 against hippocampal volume**

```{r}
main_data <- main_data %>% mutate(total_hippo=(LeftHippocampus+rightHippocampusVol))
plotRBP4vsHippovol <- main_data %>% ggplot(aes(RBP4_log10,total_hippo))+geom_point()+stat_smooth(method='lm') + theme_bw()
plotRBP4vsHippovol
```
**Figure 2** [RBP4]^ug.ml-1^ and Total Hippocampal Volume ml^-1^

**Plot RBP4 concentration against Digit Symbol**
```{r}
plotRBP4vsdigit <-  use_labels(main_data,{ggplot(..data,aes(RBP4_log10,digsym))+geom_point()+stat_smooth(method='lm') + theme_bw()})

plotRBP4vsdigit
```
**Corrlation Matrix of RBP4 against key variables**
```{r}
corrdata <- main_data %>%  filter(!is.na(eTIV)) %>% select(-c(id,`RBP4 (ug/ml)`,Sex,`Height(cm)`,`Weight(kg)`,CRP,`GripStrength(kg)`,SD,CortexVol,eTIV,BrainSegNotVent,mrtote,mrtottime)) 
corr1 <- rcorr(as.matrix(corrdata))
M <- corr1$r
M
colnames(M) <- c("Left Hippo", "Right Hippo", "Age", "ST Memory", "LT Memory", "Digit Symbol", "Verbal Fluency", "Vocabulary", "Matrix Reasoning", "Total Hippo", "Hippo/Brain", "Hippo/Skull","Brain/Skull", "lh/brain", "rh/brain","lh/skull","rh/skull", "Log10[RBP4]", "BMI")
rownames(M) <-  c("Left Hippo", "Right Hippo", "Age", "ST Memory", "LT Memory", "Digit Symbol", "Verbal Fluency", "Vocabulary", "Matrix Reasoning", "Total Hippo", "Hippo/Brain", "Hippo/Skull","Brain/Skull", "lh/brain", "rh/brain","lh/skull","rh/skull", "Log10[RBP4]", "BMI")
p_mat <- corr1$P
corr <- corrplot(M, type = "upper",method="ellipse", order = "hclust", 
         p.mat = p_mat, sig.level = 0.05, insig = "blank")
```


**There is no correlation between RBP4 concentrations and brain/cognition variables. (uncorrected)**


* Split analysis by gender
  + Female
```{r}
corrdata <- main_data %>%filter(Sex==0) %>%  filter(!is.na(eTIV)) %>% select(-c(id,`RBP4 (ug/ml)`,Sex,`Height(cm)`,`Weight(kg)`,CRP,`GripStrength(kg)`,SD,CortexVol,eTIV,BrainSegNotVent,mrtote,mrtottime)) 
corr1 <- rcorr(as.matrix(corrdata))
M <- corr1$r
#M
colnames(M) <- c("Left Hippo", "Right Hippo", "Age", "ST Memory", "LT Memory", "Digit Symbol", "Verbal Fluency", "Vocabulary", "Matrix Reasoning", "Total Hippo", "Hippo/Brain", "Hippo/Skull","Brain/Skull", "lh/brain", "rh/brain","lh/skull","rh/skull", "Log10[RBP4]", "BMI")
rownames(M) <-  c("Left Hippo", "Right Hippo", "Age", "ST Memory", "LT Memory", "Digit Symbol", "Verbal Fluency", "Vocabulary", "Matrix Reasoning", "Total Hippo", "Hippo/Brain", "Hippo/Skull","Brain/Skull", "lh/brain", "rh/brain","lh/skull","rh/skull", "Log10[RBP4]", "BMI")
p_mat <- corr1$P
corr <- corrplot(M, type = "upper",method="ellipse", order = "hclust", 
         p.mat = p_mat, sig.level = 0.05, insig = "blank")
```


*Male
```{r}
corrdata <- main_data %>%filter(Sex==1) %>%  filter(!is.na(eTIV)) %>% select(-c(id,`RBP4 (ug/ml)`,Sex,`Height(cm)`,`Weight(kg)`,CRP,`GripStrength(kg)`,SD,CortexVol,eTIV,BrainSegNotVent,mrtote,mrtottime)) 
corr1 <- rcorr(as.matrix(corrdata))
M <- corr1$r
#M
colnames(M) <- c("Left Hippo", "Right Hippo", "Age", "ST Memory", "LT Memory", "Digit Symbol", "Verbal Fluency", "Vocabulary", "Matrix Reasoning", "Total Hippo", "Hippo/Brain", "Hippo/Skull","Brain/Skull", "lh/brain", "rh/brain","lh/skull","rh/skull", "Log10[RBP4]", "BMI")
rownames(M) <-  c("Left Hippo", "Right Hippo", "Age", "ST Memory", "LT Memory", "Digit Symbol", "Verbal Fluency", "Vocabulary", "Matrix Reasoning", "Total Hippo", "Hippo/Brain", "Hippo/Skull","Brain/Skull", "lh/brain", "rh/brain","lh/skull","rh/skull", "Log10[RBP4]", "BMI")
p_mat <- corr1$P
corr <- corrplot(M, type = "upper",method="ellipse", order = "hclust", 
         p.mat = p_mat, sig.level = 0.05, insig = "blank")
```

## Regression Analysis [RBP4] and it's risk factors/ consequences

### RBP4 and the brain

RBP4 and hippocampal volume
```{r}
lr_rbp4_BV <- lm(main_data$RBP4_log10~hippo_brain,data=main_data)
main_data %>%use_labels(lm(RBP4_log10 ~ hippo_icv)) %>%summary()

```




